Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-03-06 ms.tex
recovery

\section{Model}
\label{sec:model}
\subsection{Galaxy models}
\label{sec:gal_model}
\citet{LauerFaber+:2007a} use Hubble Space Telescope WFPC2 imaging to measure the radial surface brightness profiles for hundreds of nearby early type galaxies. The measured brightness profile is generally well-fit by a ``Nuker" law parameterization:
\begin{equation}
  I(\xi)=I_b 2^{(\beta-\Gamma)/\alpha} \xi^{-\Gamma} (1+\xi^\alpha)^{-(\beta-\Gamma)/\alpha}, \,\,\,\xi\equiv\frac{r}{r_b},
\end{equation}
i.e., a broken power law that transitions from an inner power law slope, $\Gamma$, to an outer power law slope, $\beta > \Gamma$, at a break radius, $\rb$.  Assuming the stellar population is spherically symmetric, this corresponds to a stellar density $\rhostar \propto r^{-1-\Gamma}$ for $r \ll \rb$ and $\rhostar\propto r^{-1-\beta}$ for $r \gg \rb$.  To good approximation
\begin{align}
\rho_\star = 
\begin{cases}
\rho_\star|_{\rsoi} \left(r/\rsoi\right)^{-1-\Gamma} & r \leq r_b\\
\rho_\star|_{r_b} \left(r/r_b\right)^{-1-\beta} & r > r_b,
\label{eq:rhostar}
\end{cases}
\end{align}
where $\rho_\star|_{\rsoi}$ is the stellar density at the radius of the black hole sphere of influence (soi), 
\be
\rsoi \simeq G \Mbh/\sigma^2 \approx 14 M_{\bullet,8}^{0.6}\,{\rm pc},
\label{eq:rsoi}
\ee
where $M_{\bullet,8} \equiv M_{\bullet}/10^{8}M_{\odot}$ and the second equality in (\ref{eq:rsoi}) employs the $\Mbh-\sigma$ relationship of \citet{Gultekin+09},
 \begin{align}
M_{\bullet} \simeq 2\times 10^{8}\left(\frac{\sigma}{200\,{\rm
      km\,s^{-1}}}\right)^{5.1}M_{\odot}.
\label{eq:Msigma}
\end{align}
Though of questionable validity for low mass black holes (e.g., \citealt{Greene&Ho07}), we adopt equation (\ref{eq:Msigma}) for lack of a better alternative.

@@

To isolate the physics of interest, our baseline calculations neglects three potentially important effects: heat conduction, radiative cooling, and rotation.  Electron heat conduction results in an an additional term on the right hand side of the entropy equation of the form:  
\be \dot{q}_{\rm cond} = \nabla\cdot(\kappa \nabla T),
\label{eq:qdotcond}
 \ee
where $\kappa_{\rm
  eff}=\kappa_{\rm spitz}/(1+\psi)$ (\citealt{DaltonBalbus:1993a}) is
the conducitivity and $\kappa_{\rm spitz} = \kappa_0 T^{5/2}$ is the
classical \citet{Spitzer62} value ($\kappa_0\simeq 2\times 10^{-6}$ in
cgs units).  The flux limitor $\psi = \kappa_{\rm spitz} \nabla
T/(5\phi \rho c_s^3)$ saturates the conductive flux if the mean free
path for electron coulomb scattering exceeds the
temperature length scale, where $c_s \equiv (kT/\mu m_p)^{1/2}$ is the
isothermal sound speed and $\phi \lesssim 1$ is an uncertain
dimensionless constant (we adopt $\phi = 0.1$).  Even a weak magnetic
field that is oriented perpendicular to the flow could suppress the
conductivity by greatly reducing the electron mean free path.  However, for radially-decreasing temperature profiles
of interest, the flow is susceptible to the magneto-thermal instability (\citealt{Balbus01}), the non-linear evolution of which is believed to result in a radially-directed field geometry (\citealt{Parrish&Stone07}).  We show later in $\S\ref{sec:conductivity}$ that neglecting conductivity results in at most only order-unity errors in the key properties of the solutions.

Radiative cooling contributes an additional term to equation (\ref{eq:dsdt}) of the form
\be
\dot{q}_{\rm rad} = -\Lambda(T)n^{2},
\label{eq:qdot_rad}
\ee
where $n = \rho/\mu m_p$ is the gas number density and $\Lambda(T)$ is the cooling function.  We neglect radiative cooling in our baseline calculations, despite the fact that this is not justified when the wind heating $\vwO$ is low or if the mass return rate $\eta$ is high.  However, once radiative cooling becomes important compared to other sources of heating and cooling, its presence leads to thermal instability (e.g.~\citealt{McCourt+12}; $\S\ref{sec:instability}$) that cannot be accounted for by our 1D
time-independent model.  A main goal of this work is to use solutions neglecting radiation to ascertain over what range of physical conditions cooling instabilities are most likely to develop ($\S\ref{sec:instability}$).

@@

To within a constant of order unity, equation (\ref{eq:cooling2}) also
equals the ratio of the gas cooling timescale $t_{\rm cool} \equiv (3n
kT/2)/\dot{q}_{\rm rad}$ to the free-fall time $t_{\rm ff}$ at the
stagnation radius, as can be derived using the equality

@@

\citet{EmsellemCappellari+:2007a} use two-dimensional kinematic data to measure the ratio of ordered to random motion in a sample of early type galaxies, which they quantify at each galactic radius $R$ by the parameter
  \begin{align}
    \lambda_R \equiv \frac{\langle r|V|\rangle}{\langle R\sqrt{V^2+\sigma^2}\rangle} \underset{R \ll r_{\rm soi}}\sim \frac{V_{\phi}}{\sigma},
  \end{align}
where $\sigma$ is the velocity dispersion and the brackets indicate a luminosity-weighted average.  The circularization radius of the accretion flow can be written in terms of $\lambda_R$ as
\be
\frac{r_{\rm circ}}{r_{s}} \approx \frac{r_{s}}{r_{\rm soi}}\lambda_{R}^{2} \lesssim \lambda_{R}^{2},
\label{eq:rcirc}
\ee
where we have used the definition $r_{\rm soi} \equiv GM_{\bullet}/\sigma^{2}$ and in the second inequality have assumed that $r_{\rm s} \lesssim r_{\rm soi}$, a condition which is satisfied for the thermally-stable solutions of interest.

@@

Figure~\ref{fig:stag} shows our results for the stagnation radius
$r_{\rm s}$, expressed in ratio to sphere of influence radius $r_{\rm
  soi}$, with different colors showing different values of $v_{w}$.
Cusp and core galaxies are marked with square and triangles,
respectively.  Shown for comparison are our analytic results
(eq.~[\ref{eq:rs2main}]) with solid and dashed lines for cusp and core
galaxies, respectively, calculated assuming $\densSlope = 1$ and
$\densSlope= 0.5$, respectively.  Our analytic estimates are found to
provide accurate fits to the numerical results for $v_{\rm w} \gtrsim
\sigma$, but the assumptions leading to these estimates become invalid
for low $v_{\rm w}$ as the stagnation radius diverges to large radii.
For $v_{w} \gg \sigma$ the ratio $\rs/\rsoi$ does not depend
sensitively on break radius $r_{b}$, but the latter has a greater
influence when $v_{w} \lesssim \sigma$.  The three green squares which
are vertically aligned at the same value of $v_{w}/\sigma$ (close to
where the analytic estimate diverges) correspond to models with the
same black hole mass ($\Mbh = 10^7 \Msun$) and heating rate ($v_w =
300$ km s$^{-1}$) but different values of the break radius: $r_{b} = $
50, 100, and 400 pc (from top to bottom).  As $r_{\rm b} \rightarrow
r_{s}$, the effects of the break radius is to the push the stagnation
radius inwards, decreasing the SMBH accretion rate.

@@

Using the results of \citet{Bromberg+11}, the time required for a jet of luminosity $L_{\rm
  j}$ and half opening angle $\theta_{\rm j} = 0.1$ (characteristic of
AGN jets) to propagate through a gaseous mass $M_{\rm g}$ of radius $r$ is estimated to be
\begin{eqnarray}
t_{\rm jet} &\sim& 4000\,{\rm yr}\left(\frac{{L}_{\rm j}}{10^{40}\,\rm erg\,s^{-1}}\right)^{-1/3}\left(\frac{r}{\rm pc}\right)^{2/3}\left(\frac{M_{\rm g}}{10^{8}M_{\odot}}\right)^{1/3} 
\end{eqnarray}
Approximating $M_{\rm g} \sim \dot{M}t_{\rm ff}$, the ratio of the jet escape timescale to the dynamical timescale $t_{\rm dyn} \sim r/\sigma$ is given by
\be
\frac{t_{\rm jet}}{t_{\rm dyn}} \sim 7\times 10^{-3}M_{\bullet,8}^{0.13}\left(\frac{\epsilon_{\rm j}}{10^{-6}}\right)^{-1/3},
\ee
independent of $r$.  A jet with power sufficient to appreciably heat the CNM on radial scales $\sim r_{\rm s}$ ($\epsilon_{j} \gtrsim 10^{-6}$) also necessarily has sufficient power to escape the nucler region and propagate to much larger radii.  Slower outflows from the accretion disk, instead of a collimated relativistic jet, may provide a more promising source of feedback in these systems (e.g.~\citealt{Li+13}).

@@

The total gas heating rate, 
\be
\vwO = \sqrt{(v_{w}^{\star})^{2} + (v_{w}^{\rm MSP})^{2} + (v_{w}^{\rm Ia})^{2} + (v_{w}^{\bullet})^{2}},
\label{eq:vtot}
\ee
includes contributions from stellar winds, supernovae, pulsars, and radiative SMBH feedback (we neglect kinetic SMBH feedback for the moment, given its uncertain efficacy).  The strength of each heating source depends explicitly on the SMBH mass and the stellar population in the galactic nuclear region.  The latter could best be described by a single star burst formation in the past, or by a more continuous star formation history that itself varies systematically with the galaxy mass and hence $\Mbh$.

@@

\subsubsection{Continuous Star Formation History}

A single burst of star formation does not describe the average star formation histories of most galaxies.  Qualitatively,
smaller galaxies will on average have more recent star formation, resulting in energetic young stellar winds and supernovae which dominate the energy budget.  Massive galaxies, on the other hand, will on average possess older stellar populations, with their heating rates dominated by SN Ia (on large radial scales) and SMBH feedback.  We estimate the average value of $\vwO$ as a function of SMBH mass for each heating source by calculating its value using the average cosmic star formation histories of \citet{MosterNaab+:2013a}.  Appendix \ref{app:windheat} describes how the average star formation history is used to determine the stellar wind heating $v_{\rm  w}^{\star}$ and mass return parameter $\eta$ as a function of $M_{\bullet}$ (Fig.~\ref{fig:eta}; top panel).  Star formation histories are also convolved with the Ia DTD distribution to determine the SN Ia heating.

Figure~\ref{fig:vwSources} shows $\vwO(M_{\bullet})$ from each heating source (stars, MSPs, SNe Ia, and black hole feedback) calculated for the average star formation history of galaxies containing a given black hole mass.  Again, Ia heating is suppressed in low mass galaxies because the Ia radius exceeds the nominal stagnation radius (bottom panel).  MSP heating is again found to be negligible for all black hole masses.  The younger stellar populations characterizing galaxies with $\Mbh\lsim 5\times 10^{8} \Msun$ results in stellar wind and Ia heating being dominant, with $v_{w}^{\star} \sim 1000$ km s$^{-1}$ and $v_w^{\rm Ia} \sim 500$ km s$^{-1}$, respectively.  Only for the most massive galaxies $\Mbh\gsim 5\times 10^{8} \Msun$ does the lack of younger stellar populations significantly reduces the role of stellar wind heating.   Solid and dashed lines again show black hole masses for which the flow is thermally stable and thermally unstable, respectively, with $(\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|)_{r_{s}}$ as a function of $M_{\bullet}$ shown in the bottom panel.  Strikingly, {\it we find that galactic nuclei experiencing the same average star formation history as their host galaxies produce thermally stable flows across most black hole masses}.

>>>>

\section{Model}
\label{sec:model}
\subsection{Galaxy models}
\label{sec:gal_model}
\citet{LauerFaber+:2007a} use Hubble Space Telescope WFPC2 imaging to measure the radial surface brightness profiles for hundreds of nearby early type galaxies. The measured brightness profile is generally well-fit by a ``Nuker" law parameterization:
\begin{equation}
  I(\xi)=I_b 2^{(\beta-\Gamma)/\alpha} \xi^{-\Gamma} (1+\xi^\alpha)^{-(\beta-\Gamma)/\alpha}, \,\,\,\xi\equiv\frac{r}{r_b},
\end{equation}
i.e., a broken power law that transitions from an inner power law slope, $\Gamma$, to an outer power law slope, $\beta > \Gamma$, at a break radius, $\rb$.  Assuming the stellar population is spherically symmetric, this corresponds to a stellar density $\rhostar \propto r^{-1-\Gamma}$ for $r \ll \rb$ and $\rhostar\propto r^{-1-\beta}$ for $r \gg \rb$.  To good approximation
\begin{align}
\rho_\star = 
\begin{cases}
\rho_\star|_{\rsoi} \left(r/\rsoi\right)^{-1-\Gamma} & r \leq r_b\\
\rho_\star|_{r_b} \left(r/r_b\right)^{-1-\beta} & r > r_b,
\label{eq:rhostar}
\end{cases}
\end{align}
where $\rho_\star|_{\rsoi}$ is the stellar density at the radius of the black hole sphere of influence (soi), 
\begin{equation}
\rsoi \simeq G \Mbh/\sigma^2 \approx 14 M_{\bullet,8}^{0.6}\,{\rm pc},
\label{eq:rsoi}
\end{equation}
where $M_{\bullet,8} \equiv M_{\bullet}/10^{8}M_{\odot}$ and the second equality in (\ref{eq:rsoi}) employs the $\Mbh-\sigma$ relationship of \citet{Gultekin+09},
 \begin{align}
M_{\bullet} \simeq 2\times 10^{8}\left(\frac{\sigma}{200\,{\rm
      km\,s^{-1}}}\right)^{5.1}M_{\odot}.
\label{eq:Msigma}
\end{align}
Though of questionable validity for low mass black holes (e.g., \citealt{Greene&Ho07}), we adopt equation (\ref{eq:Msigma}) for lack of a better alternative.

@@

To isolate the physics of interest, our baseline calculations neglects three potentially important effects: heat conduction, radiative cooling, and rotation.  Electron heat conduction results in an an additional term on the right hand side of the entropy equation of the form:  
\begin{equation}
\dot{q}_{\rm cond} = \nabla\cdot(\kappa \nabla T),
\label{eq:qdotcond}
 \end{equation}
where $\kappa_{\rm
  eff}=\kappa_{\rm spitz}/(1+\psi)$ (\citealt{DaltonBalbus:1993a}) is
the conducitivity and $\kappa_{\rm spitz} = \kappa_0 T^{5/2}$ is the
classical \citet{Spitzer62} value ($\kappa_0\simeq 2\times 10^{-6}$ in
cgs units).  The flux limitor $\psi = \kappa_{\rm spitz} \nabla
T/(5\phi \rho c_s^3)$ saturates the conductive flux if the mean free
path for electron coulomb scattering exceeds the
temperature length scale, where $c_s \equiv (kT/\mu m_p)^{1/2}$ is the
isothermal sound speed and $\phi \lesssim 1$ is an uncertain
dimensionless constant (we adopt $\phi = 0.1$).  Even a weak magnetic
field that is oriented perpendicular to the flow could suppress the
conductivity by greatly reducing the electron mean free path.  However, for radially-decreasing temperature profiles
of interest, the flow is susceptible to the magneto-thermal instability (\citealt{Balbus01}), the non-linear evolution of which is believed to result in a radially-directed field geometry (\citealt{Parrish&Stone07}).  We show later in $\S\ref{sec:conductivity}$ that neglecting conductivity results in at most only order-unity errors in the key properties of the solutions.

Radiative cooling contributes an additional term to equation (\ref{eq:dsdt}) of the form
\begin{equation}
\dot{q}_{\rm rad} = -\Lambda(T)n^{2},
\label{eq:qdot_rad}
\end{equation}
where $n = \rho/\mu m_p$ is the gas number density and $\Lambda(T)$ is the cooling function.  We neglect radiative cooling in our baseline calculations, despite the fact that this is not justified when the wind heating $\vwO$ is low or if the mass return rate $\eta$ is high.  However, once radiative cooling becomes important compared to other sources of heating and cooling, its presence leads to thermal instability (e.g.~\citealt{McCourt+12}; $\S\ref{sec:instability}$) that cannot be accounted for by our 1D
time-independent model.  A main goal of this work is to use solutions neglecting radiation to ascertain over what range of physical conditions cooling instabilities are most likely to develop ($\S\ref{sec:instability}$).

@@

To within a constant of order unity, equation (\ref{eq:cooling2}) also
equals the ratio of the gas cooling timescale $t_{\rm cool} \equiv (3n
kT/2 \mu)/\dot{q}_{\rm rad}$ to the free-fall time $t_{\rm ff}$ at the
stagnation radius, as can be derived using the equality

@@

\citet{EmsellemCappellari+:2007a} use two-dimensional kinematic data to measure the ratio of ordered to random motion in a sample of early type galaxies, which they quantify at each galactic radius $R$ by the parameter
  \begin{align}
    \lambda_R \equiv \frac{\langle r|V|\rangle}{\langle R\sqrt{V^2+\sigma^2}\rangle} \underset{R \ll r_{\rm soi}}\sim \frac{V_{\phi}}{\sigma},
  \end{align}
where $\sigma$ is the velocity dispersion and the brackets indicate a luminosity-weighted average.  The circularization radius of the accretion flow can be written in terms of $\lambda_R$ as
\begin{equation}
\frac{r_{\rm circ}}{r_{s}} \approx \frac{r_{s}}{r_{\rm soi}}\lambda_{R}^{2} \lesssim \lambda_{R}^{2},
\label{eq:rcirc}
\end{equation}
where we have used the definition $r_{\rm soi} \equiv GM_{\bullet}/\sigma^{2}$ and in the second inequality have assumed that $r_{\rm s} \lesssim r_{\rm soi}$, a condition which is satisfied for the thermally-stable solutions of interest.

@@

Figure~\ref{fig:stag} shows our results for the stagnation radius
$r_{\rm s}$, expressed in ratio to sphere of influence radius $r_{\rm
  soi}$, with different colors showing different values of $v_{w}$.
Cusp and core galaxies are marked with square and triangles,
respectively.  Shown for comparison are our analytic results
(eq.~[\ref{eq:rs2main}]) with solid and dashed lines for cusp and core
galaxies, respectively, calculated assuming $\densSlope = 1$ and
$\densSlope= 0.6$, respectively.  Our analytic estimates are found to
provide accurate fits to the numerical results for $v_{\rm w} \gtrsim
\sigma$, but the assumptions leading to these estimates become invalid
for low $v_{\rm w}$ as the stagnation radius diverges to large radii.
For $v_{w} \gg \sigma$ the ratio $\rs/\rsoi$ does not depend
sensitively on break radius $r_{b}$, but the latter has a greater
influence when $v_{w} \lesssim \sigma$.  The three green squares which
are vertically aligned at the same value of $v_{w}/\sigma$ (close to
where the analytic estimate diverges) correspond to models with the
same black hole mass ($\Mbh = 10^7 \Msun$) and heating rate ($v_w =
300$ km s$^{-1}$) but different values of the break radius: $r_{b} = $
50, 100, and 400 pc (from top to bottom).  As $r_{\rm b} \rightarrow
r_{s}$, the effects of the break radius is to the push the stagnation
radius inwards, decreasing the SMBH accretion rate.

@@

Using the results of \citet{Bromberg+11}, the time required for a jet of luminosity $L_{\rm
  j}$ and half opening angle $\theta_{\rm j} = 0.1$ (characteristic of
AGN jets) to propagate through a gaseous mass $M_{\rm g}$ of radius $r$ is estimated to be
\begin{eqnarray}
t_{\rm jet} &\sim& 4000\,{\rm yr}\left(\frac{{L}_{\rm j}}{10^{40}\,\rm erg\,s^{-1}}\right)^{-1/3}\left(\frac{r}{\rm pc}\right)^{2/3}\left(\frac{M_{\rm g}}{10^{8}M_{\odot}}\right)^{1/3} 
\end{eqnarray}
Approximating $M_{\rm g} \sim \dot{M}t_{\rm ff}$, the ratio of the jet escape timescale to the dynamical timescale $t_{\rm dyn} \sim r/\sigma$ is given by
\begin{equation}
\frac{t_{\rm jet}}{t_{\rm dyn}} \sim 7\times 10^{-3}M_{\bullet,8}^{0.13}\left(\frac{\epsilon_{\rm j}}{10^{-6}}\right)^{-1/3},
\end{equation}
independent of $r$.  A jet with power sufficient to appreciably heat the CNM on radial scales $\sim r_{\rm s}$ ($\epsilon_{j} \gtrsim 10^{-6}$) also necessarily has sufficient power to escape the nucler region and propagate to much larger radii.  Slower outflows from the accretion disk, instead of a collimated relativistic jet, may provide a more promising source of feedback in these systems (e.g.~\citealt{Li+13}).

@@

The total gas heating rate, 
\begin{equation}
\vwO = \sqrt{(v_{w}^{\star})^{2} + (v_{w}^{\rm MSP})^{2} + (v_{w}^{\rm Ia})^{2} + (v_{w}^{\bullet})^{2}},
\label{eq:vtot}
\end{equation}
includes contributions from stellar winds, supernovae, pulsars, and radiative SMBH feedback (we neglect kinetic SMBH feedback for the moment, given its uncertain efficacy).  The strength of each heating source depends explicitly on the SMBH mass and the stellar population in the galactic nuclear region.  The latter could best be described by a single star burst formation in the past, or by a more continuous star formation history that itself varies systematically with the galaxy mass and hence $\Mbh$.

@@

\subsubsection{Continuous Star Formation History}
A single burst of star formation does not describe the typical star
formation history of most galaxies.  Qualitatively, smaller galaxies
will on average have more recent star formation, resulting in
energetic young stellar winds and supernovae which dominate the gas
heating budget.  Massive galaxies, on the other hand, will on average
possess older stellar populations, with their heating rates dominated
by SN Ia (on large radial scales) and SMBH feedback.  We estimate the
average value of $\vwO$ as a function of SMBH mass for each heating
source by calculating its value using the average cosmic star
formation histories of \citet{MosterNaab+:2013a}.  Appendix
\ref{app:windheat} describes how the average star formation history is
used to determine the stellar wind heating $v_{\rm w}^{\star}$ and
mass return parameter $\eta$ as a function of $M_{\bullet}$
(Fig.~\ref{fig:eta}; top panel).  Star formation histories are also
convolved with the Ia DTD distribution to determine the SN Ia heating.

Figure~\ref{fig:vwSources} shows $\vwO(M_{\bullet})$ from each heating
source (stars, MSPs, SNe Ia, and black hole feedback) calculated for
the average star formation history of galaxies containing a given
black hole mass.  Again, Ia heating is suppressed in low mass galaxies
because the Ia radius exceeds the nominal stagnation radius (bottom
panel).  MSP heating is again found to be negligible for all black
hole masses.  The younger stellar populations characterizing galaxies
with $\Mbh\lsim 5\times 10^{8} \Msun$ are dominated by stellar wind
and Ia heating, with $v_{w}^{\star} \sim 1000$ km s$^{-1}$ and
$v_w^{\rm Ia} \sim 500$ km s$^{-1}$, respectively.  Only for the most
massive galaxies, with $\Mbh\gsim 5\times 10^{8} \Msun$, does the lack
of young stellar populations significantly reduce the role of stellar
wind
heating.  %Solid and dashed lines again show black hole masses for which the flow is thermally stable and thermally unstable, respectively, with $(\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|)_{r_{s}}$ as a function of $M_{\bullet}$ shown in the bottom panel.
Strikingly, {\it we find that galactic nuclei experiencing the same
  average star formation history as their host galaxies produce
  thermally stable flows across all black hole masses}.  {\bf NCS:
  deleted a sentence referring to old results on thermal instability -
  it is now correct to say we are thermally stable everywhere, yes? AG
 though note our kludgy way of including Ias if rs is inside of rIa.}
